Extended Data

### 2021-10-07 doc/manuscript/Dow_manuscript_Nature.Rmd
word count
edits to intro and discussion

*NOTE: In general, this needs to be condensed. The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at ~3116 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently ok there)*

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance. 
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
However, recent findings that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that <!--leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] (lempereur_growth_2015 studied an evergreen)--> sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] suggest that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).
Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018; ] or inducing the cessation of stem growth when a critical water deficit is reached[@lempereur_growth_2015]. 
<!-- lempereur_growth_2015 studied an evergreen oak-->

@@

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
Linear models were run to examine the relationship between spring, summer and an interaction effect with ring-width for each of our 208 chronologies. The results indicated that, across all latitudes, the interacting effect of warmer springs and summers resulted in negligible increase of ring-width (supp (?) rw_drought_plot).  
*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019; @zhang_drought_2021]. (Is this relevent?)*(Neil suggested that the zhang paper would be good here but i'm not sure I see how...)
*(ADD MORE ON DROUGHT HERE)*

@@

@@

These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.

>>>>

*NOTE: In general, this needs to be condensed. The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at ~2922 words (1613 intro + results, discussion 1319--too long!) and 1 modest + 2 large display items. Articles typically have 30-50 references (currently ok there)*

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect  tree stem growth, which sequesters carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two forests, warmer spring temperatures consistently shifted the woody growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.
The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.
These findings imply that any extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to long-lived woody biomass, where it could have a substantial and lasting impact on the forest C balance. 
Thus, in contrast to global carbon cycle model expectations[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014], warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

In temperate deciduous forests, spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink[@keenan_net_2014].
The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, _NEE_)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and this expectation is represented in models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent findings that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that <!--leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] (lempereur_growth_2015 studied an evergreen)--> sink factors can limit annual growth[@zani_increased_2020; @zohner_how_2021] suggest that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.

@@

<!--1309 words before Krista's edits. Goal: 700-800-->

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`).

The observed advance in the timing of stem growth milestones under warmer springs parallels phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possible mechanisms.
First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) would lead to an earlier depletion of resources needed for growth resulting in an earlier end to the season of rapid growth ($DOY_{75}$).
Warm springs may also be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018].

@@

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`, `r EDT_tree_ring_summary`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
*It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019; @zhang_drought_2021]. (Is this relevent?)*<!-- (Neil suggested that the zhang paper would be good here but i'm not sure I see how...)-->
<!-- *(ADD MORE ON DROUGHT HERE)*-->

@@

*There may be a shift towards greater belowground allocation[@zohner_how_2021].*

@@

These findings align with those of previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019; @zhang_drought_2021] and with observations that warm springs result in earlier leaf senescence of deciduous trees because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].
They contrast, however, with model expectations[@zani_increased_2020;@buermann_widespread_2018;@keenan_net_2014] that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
